### 2022-07-21 paper/localpsf.tex
some paper updates

\begin{figure}
	\begin{center}
		\includegraphics[scale=1.0]{mean_displacement_invariance.pdf}
	\end{center}
	\caption{An operator is locally mean-displacement invariant if $\impulseresponse_{x}(y) \approx \impulseresponse_{x'}\left(y - \spatialmean(x) + \spatialmean(x')\right)$ when $x$ is close to $x'$. Here $\spatialmean(x)$ is the mean (center of mass) of $\phi_x$, and similar for $\spatialmean(x')$.}
	\label{fig:mean_displacement_invariance}
\end{figure}

@@

\begin{figure}[H]
	\begin{subfigure}{0.275\textwidth}
    \includegraphics[scale=0.10]{meshHeight_view0_edges_withColorBar.png}	
    \end{subfigure}
    \begin{subfigure}{0.275\textwidth}
    \includegraphics[scale=0.10]{trueVelocity_glyphs_withColorBar.png}
    \end{subfigure}
    \begin{subfigure}{0.275\textwidth}
    \includegraphics[scale=0.135]{mtrue_withColorBar.png}
    \end{subfigure}
\end{figure}

@@

\section{Ellipsoid intersection test}
\label{sec:fast_ellipsoid_intersection_test}
The procedure for choosing sample points relies on quickly determining whether two ellipsoids intersect. Let $E_1$ and $E_2$ be the ellipsoids defined as
\begin{align*}
	E_1 :=& \{x : (x - \spatialmean_1)^T \spatialcov_1^{-1} (x - \spatialmean_1) \le \tau^2\}, \\
	E_2 :=& \{x : (x - \spatialmean_2)^T \spatialcov_2^{-1} (x - \spatialmean_2) \le \tau^2\}, \\
\end{align*}
where $\spatialmean_1, \spatialmean_2 \in \mathbb{R}^\gdim$, and $\spatialcov_1, \spatialcov_2 \in \mathbb{R}^{\gdim \times \gdim}$ are positive definite. Let $K$ be the following one dimensional convex function:
\begin{equation*}
	K(s) := 1 - \frac{1}{\tau^2} (\spatialmean_1 - \spatialmean_2)^T \left(\frac{1}{1-s}\spatialcov_1 + \frac{1}{s}\spatialcov_2\right)^{-1}(\spatialmean_1 - \spatialmean_2)	
\end{equation*}
In \cite{GilitschenskiHanebeck12} it is shown that $E_1 \cap E_2 = \{\}$ if and only if $K(s) < 0$ for some $s\in (0,1)$. We check whether $E_1$ and $E_2$ intersect by minimizing $K(s)$ on $(0,1)$. If $K(s^*) <0$ at the minimizer $s^*$, then $E_1 \cap E_2 = \{\}$. Otherwise $E_1 \cap E_2 \neq \{\}$.

\begin{figure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.40]{ellipsoids_intersect0.pdf}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.40]{ellipsoids_intersect1.pdf}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.40]{ellipsoids_intersect2.pdf}
	\end{subfigure}
	\caption{The ellipsoids intersect if $K(s) \ge 0$ for all $s \in (0,1)$.
	}
	\label{fig:ellipsoid_intersection_test}
\end{figure}

The function $K(s)$ may be evaluated quickly for many $s$ by pre-computing the solution to the generalized eigenvalue problem
\begin{equation*}
	\spatialcov_1 P = \spatialcov_2 P \Lambda,
\end{equation*}
where $P \in \mathbb{R}^{\gdim \times \gdim}$ is the matrix of generalized eigenvectors (which may be non-orthogonal), and $\Lambda = \diag(\lambda_1,\lambda_2,\dots,\lambda_\gdim)$ is the diagonal matrix of generalized eigenvalues $\lambda_i$. The matrix $P$ simultaneously diagonalizes $\spatialcov_1$ and $\spatialcov_2$, in the sense that $P^T\spatialcov_1 P = \Lambda$, and $P^T\spatialcov_2 P = I$, where $I$ is the $\gdim \times \gdim$ identity matrix. Using this diagonalization, and some algebraic manipulations, we may write $K(s)$ as
\begin{equation}
	\label{eq:Ks_generalized}
	K(s) = 1 - \frac{1}{\tau^2} \sum_{i=1}^\gdim \frac{s(1-s)}{1 + s (\lambda_i - 1)}v_i^2,
\end{equation}
where $v := P^T\left(\spatialmean_1 - \spatialmean_2\right)$. We compute the generalized eigenvalue decomposition of $\spatialcov_1$ and $\spatialcov_2$, then minimize $K(s)$ in the form \eqref{eq:Ks_generalized} on the interval $(0,1)$ using Brent's algorithm \cite{Brent71} (any fast 1 dimensional convex optimization routine may be used). The resulting algorithm for checking whether $E_1$ and $E_2$ intersect is summarized in Algorithm \ref{alg:ellipsoid_intersection_test}.

\begin{algorithm2e}
	\SetAlgoNoLine
	\SetKwInOut{Input}{Input}
	\SetKwInOut{Output}{Output}
	\SetKwProg{Fn}{Function}{}{}
	\Input{Ellipsoid $E_1$ with mean $\spatialmean_1$ and covariance $\spatialcov_1/\tau^2$\\
		Ellipsoid $E_2$ with mean $\spatialmean_2$ and covariance $\spatialcov_2/\tau^2$ 
	}

	\Output{Boolean $\text{ellipsoids\_intersect}$ which is true if $E_1 \cap E_2 \neq \{\}$ and false otherwise}

	Solve generalized eigenvalue problem $\spatialcov_1 P = \spatialcov_2 P \Lambda$

	$v \gets P^T\left(\spatialmean_1 - \spatialmean_2\right)$

	$\displaystyle K^* \gets \min_{s \in (0,1)}~1 - \frac{1}{\tau^2} \sum_{i=1}^\gdim \frac{s(1-s)}{1 + s (\lambda_i - 1)}v_i^2$

	\If{$K^* < 0$}{

		$\text{ellipsoids\_intersect} \gets \text{False}$

	}
	\Else{

		$\text{ellipsoids\_intersect} \gets \text{True}$

	}
	\caption{Determining whether two ellipsoids intersect}
	\label{alg:ellipsoid_intersection_test}
\end{algorithm2e}

@@

\begin{figure}
	\centering
	\includegraphics[scale=0.5]{spd_rational_function.pdf}
	\caption{Rational function $\Pi_\ratord(\lambda)$ for $a=-1$, $b=0$, and $\ratord \in \{1,2,3\}$. As $\ratord$ increases, $\Pi_\ratord$ approaches the indicator function for the set $(a,b)$. }
	\label{fig:spd_rat}
\end{figure}

Computing $\mathbf{\Pi}=\Pi(\mathbf{A}_H^{\text{sym}})$ directly is difficult, but we may approximate $\Pi$ by a high order rational function $\Pi_\ratord \approx \Pi$ which is defined as follows:
\begin{equation*}
	 \Pi_\ratord(\lambda) := \frac{1}{1 + T(\lambda)^{2^\ratord} }, \qquad \text{where} \qquad T(\lambda) := \frac{2 \lambda - (b+a)}{b-a}.
\end{equation*}
Here $\ratord$ is a positive integer; the larger $\ratord$ is, the better the approximation $\Pi \approx \Pi_\ratord$. This is illustrated in Figure \ref{fig:spd_rat}. Replacing scalar operations with matrix operations, we have an approximation $\mathbf{\Pi}_\ratord \approx \mathbf{\Pi}$ given by
\begin{equation}
\label{eq:rational343}
\mathbf{\Pi}_\ratord := \Pi_\ratord\left(\mathbf{A}_H^{\text{sym}}\right) = \left(\mathbf{I} + \mathbf{T}^{2^\ratord}\right)^{-1},
\end{equation}
where
\begin{equation*}
\mathbf{T} := T(\mathbf{A}_H^{\text{sym}}) = \left(2 \mathbf{A}_H^{\text{sym}} - (b+a) \mathbf{I}\right)/(b-a).
\end{equation*}
Using $\mathbf{\Pi}_\ratord$ instead of $\mathbf{\Pi}$ in \eqref{eq:I_Pi_abs} allows us to efficiently approximate $|\mathbf{A}_H^{\text{sym}}|$ using H-matrix methods. 
This approximation procedure is as follows. First, we compute $\lambda_\text{min}$ using the Lanczos method, and set $a=\gamma \lambda_\text{min}$ and $b=0$, where $\gamma\ge 1$ is a parameter that must be chosen. 
The method is relatively insensitive to $\gamma$ and numerically we find that the method works well for any $\gamma \in [1,2]$. We use $\gamma=1.5$ in our numerical results. 
Next, we compute $\mathbf{T}$ using fast H-matrix scalar multiplication and addition. We form the $\mathbf{T}^{2^\ratord}$ via repeated squaring, $\mathbf{T}^2 = \mathbf{T} \mathbf{T}$, $\mathbf{T}^4 = \mathbf{T}^2 \mathbf{T}^2$, $\mathbf{T}^8 = \mathbf{T}^4 \mathbf{T}^4$ and so on, where each of these matrix squarings is performed using H-matrix multiplication, which is fast. We recommend using $\ratord=1$ or $2$ (we use $\ratord=1$ in our numerical results). Larger $\ratord$ is unnecessary and may make the method unstable. The approximate spectral projector $\mathbf{\Pi}_\ratord$ is formed from $\mathbf{T}^{2^\ratord}$ per \eqref{eq:rational343}, using fast H-matrix operations to perform the required matrix-matrix addition and matrix inversion. Finally, a positive semi-definite approximation $\mathbf{A}_H^{\text{sym}+}$ of $\mathbf{A}_H^{\text{sym}}$ is formed as
\begin{equation*}
\mathbf{A}_H^{\text{sym}+} := \left(\mathbf{I} - 2\mathbf{\Pi}_\ratord\right)\mathbf{A}_H^{\text{sym}},
\end{equation*}
where fast H-matrix operations are used to perform the necessary matrix-matrix subtraction and multiplication. Note that for any $\ratord \ge 1$, $\Pi_\ratord(\lambda)> 1/2$ for $\lambda \in (a,b)$, so $\mathbf{A}_H^{\text{sym}+}$ is positive semi definite\footnote{Some indefiniteness may be introduced by truncation error in the H-matrix operations, but the negative eigenvalues caused by this truncation error are typically extremely small.}. At the end, $\mathbf{A}_H^{\text{sym}+}$ is symmetrized once again, to correct for any small non-symmetries that are introduced by truncation error in the H-matrix operations. Once constructed, we use the symmetric positive semidefinite matrix $\mathbf{A}_H^{\text{sym}+}$ in place of $\mathbf{A}_H$ in further linear algebra operations.

>>>>

\begin{figure}
	\begin{center}
		\includegraphics[scale=0.75]{mean_displacement_invariance.pdf}
	\end{center}
	\caption{An operator is locally mean-displacement invariant if $\impulseresponse_{x}(y) \approx \impulseresponse_{x'}\left(y - \spatialmean(x) + \spatialmean(x')\right)$ when $x$ is close to $x'$. Here $\spatialmean(x)$ is the mean (center of mass) of $\phi_x$, and similar for $\spatialmean(x')$.}
	\label{fig:mean_displacement_invariance}
\end{figure}

@@

\begin{figure}[H]
	\begin{subfigure}{0.32\textwidth}
    \includegraphics[scale=0.12]{meshHeight_view0_edges_withColorBar.png}	
    \end{subfigure}
    \begin{subfigure}{0.32\textwidth}
    \includegraphics[scale=0.12]{trueVelocity_glyphs_withColorBar.png}
    \end{subfigure}
    \begin{subfigure}{0.32\textwidth}
    \includegraphics[scale=0.162]{mtrue_withColorBar.png}
    \end{subfigure}
\end{figure}

@@

\section{Ellipsoid intersection test}
\label{sec:fast_ellipsoid_intersection_test}
The procedure for choosing sample points relies on quickly determining whether two ellipsoids intersect. Let $E_1$ and $E_2$ be the ellipsoids defined as
\begin{align*}
	E_1 :=& \{x : (x - \spatialmean_1)^T \spatialcov_1^{-1} (x - \spatialmean_1) \le \tau^2\}, \\
	E_2 :=& \{x : (x - \spatialmean_2)^T \spatialcov_2^{-1} (x - \spatialmean_2) \le \tau^2\}, \\
\end{align*}
where $\spatialmean_1, \spatialmean_2 \in \mathbb{R}^\gdim$, and $\spatialcov_1, \spatialcov_2 \in \mathbb{R}^{\gdim \times \gdim}$ are positive definite. Let $K$ be the following one dimensional convex function:
\begin{equation*}
	K(s) := 1 - \frac{1}{\tau^2} (\spatialmean_1 - \spatialmean_2)^T \left(\frac{1}{1-s}\spatialcov_1 + \frac{1}{s}\spatialcov_2\right)^{-1}(\spatialmean_1 - \spatialmean_2)	
\end{equation*}
In \cite{GilitschenskiHanebeck12} it is shown that $E_1 \cap E_2 = \{\}$ if and only if $K(s) < 0$ for some $s\in (0,1)$ (see Figure \ref{fig:ellipsoid_intersection_test}).

\begin{figure}
	\begin{subfigure}{0.37\textwidth}
		\centering
		\includegraphics[scale=1.0]{ellipsoids_intersect0.pdf}
	\end{subfigure}
	\begin{subfigure}{0.30\textwidth}
		\centering
		\includegraphics[scale=1.0]{ellipsoids_intersect1.pdf}
	\end{subfigure}
	\begin{subfigure}{0.30\textwidth}
		\centering
		\includegraphics[scale=1.0]{ellipsoids_intersect2.pdf}
	\end{subfigure}
	\caption{The ellipsoids intersect if $K(s) \ge 0$ for all $s \in (0,1)$.
	}
	\label{fig:ellipsoid_intersection_test}
\end{figure}

The function $K(s)$ may be evaluated quickly for many $s$ by pre-computing the solution to the generalized eigenvalue problem
\begin{equation*}
	\spatialcov_1 P = \spatialcov_2 P \Lambda,
\end{equation*}
where $P \in \mathbb{R}^{\gdim \times \gdim}$ is the matrix of generalized eigenvectors (which may be non-orthogonal), and $\Lambda = \diag(\lambda_1,\lambda_2,\dots,\lambda_\gdim)$ is the diagonal matrix of generalized eigenvalues $\lambda_i$. The matrix $P$ simultaneously diagonalizes $\spatialcov_1$ and $\spatialcov_2$, in the sense that $P^T\spatialcov_1 P = \Lambda$, and $P^T\spatialcov_2 P = I$, where $I$ is the $\gdim \times \gdim$ identity matrix. Using this diagonalization, and some algebraic manipulations, we may write $K(s)$ as
\begin{equation}
	\label{eq:Ks_generalized}
	K(s) = 1 - \frac{1}{\tau^2} \sum_{i=1}^\gdim \frac{s(1-s)}{1 + s (\lambda_i - 1)}v_i^2,
\end{equation}
where $v := P^T\left(\spatialmean_1 - \spatialmean_2\right)$.

To check whether $E_1$ and $E_2$ intersect, we compute the generalized eigenvalue decomposition of $\spatialcov_1$ and $\spatialcov_2$, form $v$, then minimize $K(s)$ in the form \eqref{eq:Ks_generalized} on the interval $(0,1)$ using Brent's minimization algorithm \cite{Brent71} (any fast 1 dimensional convex optimization routine may be used). If $K(s^*) <0$ at the minimizer $s^*$, then $E_1 \cap E_2 = \{\}$. Otherwise $E_1 \cap E_2 \neq \{\}$.
%The resulting algorithm for checking whether $E_1$ and $E_2$ intersect is summarized in Algorithm \ref{alg:ellipsoid_intersection_test}.

%\begin{algorithm2e}
%	\SetAlgoNoLine
%	\SetKwInOut{Input}{Input}
%	\SetKwInOut{Output}{Output}
%	\SetKwProg{Fn}{Function}{}{}
%	\Input{Ellipsoid $E_1$ with mean $\spatialmean_1$ and covariance $\spatialcov_1/\tau^2$\\
%		Ellipsoid $E_2$ with mean $\spatialmean_2$ and covariance $\spatialcov_2/\tau^2$ 
%	}
%	
%	\Output{Boolean $\text{ellipsoids\_intersect}$ which is true if $E_1 \cap E_2 \neq \{\}$ and false otherwise}
%	
%	
%	Solve generalized eigenvalue problem $\spatialcov_1 P = \spatialcov_2 P \Lambda$
%		
%	$v \gets P^T\left(\spatialmean_1 - \spatialmean_2\right)$
%		
%	$\displaystyle K^* \gets \min_{s \in (0,1)}~1 - \frac{1}{\tau^2} \sum_{i=1}^\gdim \frac{s(1-s)}{1 + s (\lambda_i - 1)}v_i^2$
%		
%	\If{$K^* < 0$}{
%			
%		$\text{ellipsoids\_intersect} \gets \text{False}$
%			
%	}
%	\Else{
%			
%		$\text{ellipsoids\_intersect} \gets \text{True}$
%			
%	}
%	\caption{Determining whether two ellipsoids intersect}
%	\label{alg:ellipsoid_intersection_test}
%\end{algorithm2e}

@@

%\begin{figure}
%	\centering
%	\includegraphics[scale=0.5]{spd_rational_function.pdf}
%	\caption{Rational function $\Pi_\ratord(\lambda)$ for $a=-1$, $b=0$, and $\ratord \in \{1,2,3\}$. As $\ratord$ increases, $\Pi_\ratord$ approaches the indicator function for the set $(a,b)$. }
%	\label{fig:spd_rat}
%\end{figure}

Computing $\mathbf{\Pi}=\Pi(\mathbf{A}_H^{\text{sym}})$ directly is difficult, but we may approximate $\Pi$ by a high order rational function $\Pi_\ratord \approx \Pi$ which is defined as follows:
\begin{equation*}
	 \Pi_\ratord(\lambda) := \frac{1}{1 + T(\lambda)^{2^\ratord} }, \qquad \text{where} \qquad T(\lambda) := \frac{2 \lambda - (b+a)}{b-a}.
\end{equation*}
Here $\ratord$ is a positive integer; the larger $\ratord$ is, the better the approximation $\Pi \approx \Pi_\ratord$. 
%This is illustrated in Figure \ref{fig:spd_rat}. 
Replacing scalar operations with matrix operations, we have an approximation $\mathbf{\Pi}_\ratord \approx \mathbf{\Pi}$ given by
\begin{equation}
\label{eq:rational343}
\mathbf{\Pi}_\ratord := \Pi_\ratord\left(\mathbf{A}_H^{\text{sym}}\right) = \left(\mathbf{I} + \mathbf{T}^{2^\ratord}\right)^{-1},
\end{equation}
where
\begin{equation*}
\mathbf{T} := T(\mathbf{A}_H^{\text{sym}}) = \left(2 \mathbf{A}_H^{\text{sym}} - (b+a) \mathbf{I}\right)/(b-a).
\end{equation*}
Using $\mathbf{\Pi}_\ratord$ instead of $\mathbf{\Pi}$ in \eqref{eq:I_Pi_abs} allows us to efficiently approximate $|\mathbf{A}_H^{\text{sym}}|$ using H-matrix methods. 
This approximation procedure is as follows. First, we compute $\lambda_\text{min}$ using the Lanczos method, and set $a=\gamma \lambda_\text{min}$ and $b=0$, where $\gamma\ge 1$ is a parameter that must be chosen. 
The method is relatively insensitive to $\gamma$ and numerically we find that the method works well for any $\gamma \in [1,2]$. We use $\gamma=1.5$ in our numerical results. 
Next, we compute $\mathbf{T}$ using fast H-matrix scalar multiplication and addition. We form the $\mathbf{T}^{2^\ratord}$ via repeated squaring, $\mathbf{T}^2 = \mathbf{T} \mathbf{T}$, $\mathbf{T}^4 = \mathbf{T}^2 \mathbf{T}^2$, $\mathbf{T}^8 = \mathbf{T}^4 \mathbf{T}^4$ and so on, where each of these matrix squarings is performed using H-matrix multiplication, which is fast. We recommend using $\ratord=1$ or $2$ (we use $\ratord=1$ in our numerical results). Larger $\ratord$ is unnecessary and may make the method unstable. The approximate spectral projector $\mathbf{\Pi}_\ratord$ is formed from $\mathbf{T}^{2^\ratord}$ per \eqref{eq:rational343}, using fast H-matrix operations to perform the required matrix-matrix addition and matrix inversion. Finally, a positive semi-definite approximation $\mathbf{A}_H^{\text{sym}+}$ of $\mathbf{A}_H^{\text{sym}}$ is formed as
\begin{equation*}
\mathbf{A}_H^{\text{sym}+} := \left(\mathbf{I} - 2\mathbf{\Pi}_\ratord\right)\mathbf{A}_H^{\text{sym}},
\end{equation*}
where fast H-matrix operations are used to perform the necessary matrix-matrix subtraction and multiplication. Note that for any $\ratord \ge 1$, $\Pi_\ratord(\lambda)> 1/2$ for $\lambda \in (a,b)$, so $\mathbf{A}_H^{\text{sym}+}$ is positive semi definite\footnote{Some indefiniteness may be introduced by truncation error in the H-matrix operations, but the negative eigenvalues caused by this truncation error are typically extremely small.}. At the end, $\mathbf{A}_H^{\text{sym}+}$ is symmetrized once again, to correct for any small non-symmetries that are introduced by truncation error in the H-matrix operations. Once constructed, we use the symmetric positive semidefinite matrix $\mathbf{A}_H^{\text{sym}+}$ in place of $\mathbf{A}_H$ in further linear algebra operations.
